Push-Pull Signal-Function Functional Reactive Programming

### 2012-11-30 Docs/IFL2012/design.tex
Performance results in discussion section - graphic not rendering properly

The purpose of FRP is to provide an efficient, declarative abstraction for creating
reactive programs. To this end, there are three goals that the TimeFlies
system is intended to meet. \rn{What are the three!?  Enumerate.  Push
based, composable, IO-system agnostic?}

Efficient evaluation is the motivation for push-based evaluation of events.
Since FRP programs are expected to  interact with an outside world in real time,
efficiency cannot simply be measured by the runtime of a program. Thus, when speaking of efficiency,
we are expressing a desire that the system utilize as few system resources as possible
for the task at hand, while responding as quickly as possible to external inputs and
producing output at a consistently high sample rate.

\todo{At this point we still need a clear notion of what a pull-based
  evaluation of events would actually do.}

@@

\rn{Can the above be made into a slightly more technical claim?
  Composable abstractions enable composing $O(N)$ components while
  reasoning about the $O(N)$ local interactions, rather than the
  $O(N^2)$ global interactions between all components?  Second, I
  think not needing to know the implementation is separate from
  ``composability''; surely there is some name for that: ADTs,
  ``encapsulation'', information hiding, etc.}

% It is fine for a system to be composable with regards to itself, but
% an FRP system must
\new{Further, an FRP program is not a closed system; it}
interacts with the outside world. Since we cannot anticipate every
possible form of input and output that the system will be asked to interact
with, we must interface with Haskell's IO system. 
\rn{First mention of Haskell... need some transition?}
In particular, most libraries
for user interaction (e.g. GUI and graphics libraries such as GTK+ and GLUT) and
most libraries for time-dependent IO (e.g. audio and video systems) make use of
the event-loop abstraction. In this abstraction, event handlers are registered
\new{centrally}; 
\new{once initiated, the running event-loop detects events and invokes handlers; and
  handlers are responsible for rendering output.}
%% and then a command is issued to run a loop which detects events
%% and runs the handlers, and uses the results of the handlers to render the
%% appropriate output.

\rn{Is the event loop really an {\em abstraction}?  What is
  abstracted?  Or is it just a ``model'' or an API?}

We would like for the FRP system to be easy to integrate with such IO systems,
while being flexible enough to enable its use with other forms of IO systems,
such as simple imperative systems, threaded systems, or network servers.
\rn{vague... not sure what these mean.  Threaded = multiple event
  loop?  Alternative to an event loop is... what?... blocking IO operations?}

@@

A rigorous 
%and formal elucidation of the 
semantics of signal-function FRP with signal
vectors remains unattempted, though a categorical semantics for signal-function
FRP using {\em fan categories} has been explored~\cite{Jeffrey2012}. In both
these semantics and their implementation, events are represented (following
AFRP~\cite{Courtney2001-1} and Yampa~\cite{Nilsson2005}) as option-valued
signals. A push-based implementation of these semantics is proposed, but the
unit being pushed is a {\em segment} of signals, and not event occurrences. This
approach requires processing the whole signal function at each time step,
whether an event occurs or not, and checking for its occurrence at each step.

\rn{Might want to make it clear -- the ``whole'' signal function means
  ``all signal functions in the program'' (i.e. which are composed
  to form the global signal function) right?}

Another concern is that this approach limits the event rate to the sampling rate.
The rate of sampling should, at some level, not matter to the FRP system. Events
which occur between sampling intervals are never observed by the
system. \rn{An FRP system has sampling intervals?  Where were we told
  this!?}  
Since the signal function is never evaluated except at the sampling intervals, no
events can otherwise be observed.

This \new{raises another} concern. Events are not instantaneous in this
formulation. If events are signals, the sampling instant must fall within
the interval where there is an event occurrence present for that event to be
observed. If events are instantaneous, the probability of observing an event 
occurrence is zero. \rn{So.... how do existing systems/semantics
  compensate for this?  They allow events to have temporal extent?}

Therefore, TimeFlies employs the N-Ary FRP type formulation to represent signals and
events as distinct entities in the inputs and outputs of signal functions. This means
we are now free to choose our representation of events, and to separate it from the
representation and evaluation of signals.
%
This freedom yields the additional ability to make events independent of the sampling
interval altogether. The semantics of event handling in TimeFlies is that an event occurrence
is responded to immediately, and does not wait for the next sampling
instant. 
\rn{Could you flesh out how this is a {\em semantics} issue and not
  merely a more efficient or accurate implementation?  Ah, that's
  answered in the next para, but might be good to use the word
  semantics somewhere and connect this explicitly.} 
This allows events to be instantaneous, and further, allows multiple
events to occur within a single sampling interval.
\rn{Clarification---even in traditional approaches multiple {\bf
    different} events could occur in one sampling round right?  It's
  just that {\em each} event [stream] could only have one occurrence per
  sampling interval?}

There are two tradeoffs intrinsic to this approach. The first is that events are only partially ordered
temporally. There is no way to guarantee the order of observation of event occurrences occurring in the
same sampling interval. Further, the precise time of an event occurrence cannot be observed, only the 
time of the last \rn{[signal?]} sample prior to the occurrence.
\rn{That last sentence had an ambiguous parse wherein the first time I read it I thought it was saying that you
couldn't tell the timestamp of event N only of the preceding event $N-1$, which of course makes no sense,
but it might be good to clarify what {\em sample} you're referring to.}

In return for giving up total ordering and precise observation of the timing of events, we obtain the
ability to employ push-based evaluation for event occurrences, and the ability to non-deterministically
merge event occurrences. When events being input to a non-deterministic merge have simultaneous occurrences,
we arbitrarily select one to occur first. This does not violate any guarantee about time values, since
they will both have the same time value in either case, and does not violate any guarantee about ordering,
since no guarantee of their order is given.
\rn{Nice.}

@@

In order to type signal functions, we must be able to describe their input and
output. In most signal function systems, a signal function takes exactly one
input signal and produces exactly one output signal. Multiple inputs or outputs are handled
by making the input or output a signal of tuples, and combinators which combine or split
the inputs or outputs of a signal assume this.  Events are represented at the
type level as a particular type of signal, and at the value level as an option:
either an event occurrence or not. 
\rn{That's a bit confusing, because of the issue of whether we're talking about {\em event occurrences} or
  event streams.  At runtime there is only a representation for the former, which is a simple {\tt Maybe},
  right?}

@@

Our interface for evaluating signal functions takes the form of a monad
transformer~\cite{Jones1995}. This formulation permits us to provide specific
temporal actions to perform on a signal function in sequence, and in conjunction
with actions taken in Haskell's {\tt IO} monad, or another monad. Actions are
provided in this monad to push an event to the signal function's input, step
the signal function's time, or modify a signal on the signal functions input to
be observed at the next time step. Outputs are handled by a vector \rn{vector? in the same sense as a
  signal vector?} of actuation
functions in the underlying monad, supplied by the user of the interface. This
vector is structured by parameterizing it over the signal function's output
signal vector. \rn{huh? more concrete way to say it?}

@@

Signal functions are constructed from combinators, which are primitive signal
functions and operations to combine these primitives. These combinators are
grouped as basic signal functions, lifting operations for pure functions,
routing, reactivity, feedback, event processing, joining, and time dependence.

The basic signal functions
provide very simple operations. The {\tt identity} signal function, as expected,
simply copies its input to its output. The {\tt constant} signal function
produces the provided value as a signal at all times. The {\tt never} signal
function has an event output which never produces occurrences. The {\tt asap}
function produces an event occurrence with the given value at the first time
step after it is switched into the network. The {\tt after} function waits for
the specified amount of time before producing an event occurrence.

Two combinators are provided to lift pure functions to signal functions.
The {\tt pureSignalTransformer} combinator applies the pure function pointwise
to a signal. The {\tt pureEventTransformer} combinator applies the function to
the value carried by each occurrence of an input event.

The routing combinators are used to combine signal functions, and to re-arrange
and discard components of signal vectors in order to connect signal functions.

Only those routing combinators which modify or combine signal functions
({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (may replace themselves in
response to events, and then only if they inherit their reactivity from the
signal function(s) which are inputs to combinator. The rest do not
react to or modify the input in any way, except to re-arrange it, copy it, or
discard it altogether.

Reactivity is introduced by means of the {\tt switch} combinator.
The design of this combinator
allows modular packaging of reactivity. A signal function can determine
autonomously when to replace itself, based only on its input and state,
by emitting an event occurrence carrying its replacement. The combinator
consumes and hides the event carrying the replacement signal function, 
so the reactivity is not exposed by the resulting reactive signal function.

It is often useful for a signal function to receive a portion of its
own output as input. This is especially useful when we have two
signal functions which we would like to mutually interact. We cannot
just serially compose them, we must also bring the output of the second
back around to the first. Many signal-processing algorithms also depend
on feedback. This ability is provided by the {\tt loop} operator.

This combinator provides decoupling for signals
(the input signal is the output signal at the previous time-step)
but not events (event occurrences are supplied to the combinator immediately).
This means that the programmer has the responsibility to ensure that feedback
does not generate an infinite sequence of events in a single time-step.

The {\tt filter} and {\tt filterList} combinators are stateless signal functions
which apply a function to each event occurrence. The {\tt filter} combinator
expects an option-typed output, while the {\tt list} combinator expects a list
output. If the function yields a non-empty output, one or more output occurrences
are produced with the elements of that output. If the output is empty, there
are no output occurrences.

The {\tt accumulate} and {\tt accumulateList} functions are similar, except that
they maintain a state value which is replaced after each input event occurrence
by an additional output from the function, and supplied to the function in
addition to the value of each event occurrence.

The joining combinators provide the ability to combine two event
streams, two signals, or a signal and an event stream.

The {\tt union} combinator is a non-deterministic merge of two event streams.
Simultaneity of events is only observable up to the sampling rate, but events
can occur in between sampling intervals. Thus, if two events occur
simultaneously, one occurrence is arbitrarily selected to be first in the output
stream.

The {\tt combineSignals} combinator applies a binary function pointwise to two
signals, and produces the result of this application as a third signal. The
combining function is necessary because we will have two input samples at each
time step, and must produce exactly one output sample.

The {\tt capture} combinator adds the last-sampled value of a signal at the time
of an event occurrence to that event occurrence.

These three combinators together provide the ability to combine elements of
signal vectors. By combining these combinators, along with the {\tt cancelLeft}
and {\tt cancelRight} routing combinators, arbitrary signal vectors can be
reduced.

A set of combinators are provided for making use of time-dependence in a signal
function. These combinators allow the modification of signals and events with
respect to time, and the observation of the current time value.

The simplest time-dependent combinator is {\tt time}, which simply outputs
the time since it began to be evaluated. This does not necessarily correspond to
the time since the global signal function began to be evaluated, since the
signal function in which the {\tt time} combinator is used may have been
introduced through {\tt switch}.

The {\tt delay} signal function allows events to be delayed. An initial delay
time is given, and event occurrences on the right input can carry a new delay
time. Event occurrences on the left input are stored and output when their delay
time has passed. Changing the delay time does not affect occurrences already
waiting.

The {\tt integrate} combinator outputs the rectangle-rule approximation of the
integral of its input signal with respect to time.

@@

{\tt SFEvalT} actions are constructed using combinators to push events,
update inputs, and step time, as well as actions lifted from the underlying
monad (used to obtain these inputs). An action in the underlying monad
which produces the result and a new state is obtained with the {\tt runSFEvalT}
function.

>>>>

The purpose of FRP is to provide an efficient, declarative abstraction for creating
reactive programs. To this end, there are three goals that the TimeFlies
system is intended to meet:

An FRP system should permit efficient evaluation. Since FRP programs are
expected to  interact with an outside world in real time, efficiency cannot
simply be measured by the runtime of a program. Thus, when speaking of
efficiency, we are expressing a desire that the system utilize as few system
resources as possible for the task at hand, while responding as quickly as
possible to external inputs and producing output at a consistently high sample
rate.

@@

Further, an FRP program is not a closed system; it interacts with the outside
world. Since we cannot anticipate every possible form of input and output that
the system will be asked to interact with, we must interface with IO systems.
In particular, most libraries for user interaction (e.g. GUI and graphics
libraries such as GTK+ and GLUT) and most libraries for time-dependent IO
(e.g. audio and video systems) make use of the event-loop model. In this
model, event handlers are registered for various forms of input, and then a
loop is initiated which checks for, or blocks waiting for, input and then
actuates the appropriate handlers to compute on the input and actuate outputs.

We would like for the FRP system to be easy to integrate with such IO systems,
while being flexible enough to enable its use with other forms of IO systems,
such as systems which use blocking IO and threads, rather than event loops.

@@

A rigorous semantics of signal-function FRP with signal vectors remains
unattempted, though a categorical semantics for signal-function FRP using
{\em fan categories} has been explored~\cite{Jeffrey2012}. In both these
semantics and their implementation, events are represented (following
AFRP~\cite{Courtney2001-1} and Yampa~\cite{Nilsson2005}) as option-valued
signals. A push-based implementation of these semantics is proposed, but the
unit being pushed is a {\em segment} of signals, and not event occurrences. This
approach requires processing the whole signal function at each time step,
whether an event occurs or not, and checking for its occurrence at each step.

Another concern is that this approach limits the event rate to the sampling rate.
The rate of sampling should, at some level, not matter to the FRP system. Events
which occur between sampling intervals are never observed by the
system. Since the signal function is never evaluated except at the sampling
intervals, no events can otherwise be observed.

This raises another concern. Events are not instantaneous in this
formulation. If events are signals, the sampling instant must fall within
the interval where there is an event occurrence present for that event to be
observed. If events are instantaneous, the probability of observing an event 
occurrence is zero. So in current signal-function semantics, events occurrences
implicitly have some duration which contains the point of sampling.

Therefore, TimeFlies employs the N-Ary FRP type formulation to represent signals and
events as distinct entities in the inputs and outputs of signal functions. This means
we are now free to choose our representation of events, and to separate it from the
representation and evaluation of signals.

This freedom yields the additional ability to make events independent of the
sampling interval altogether. The semantics of event handling in TimeFlies is
that an event occurrence is responded to immediately, and does not wait for the
next sampling instant. This allows events to be instantaneous, and further,
allows multiple occurrences for an event within a single sampling interval.

There are two tradeoffs intrinsic to this approach. The first is that events are
only partially ordered temporally. There is no way to guarantee the order of
observation of event occurrences occurring in the same sampling interval.
Further, the precise time of an event occurrence cannot be observed, only the 
time that the signal function was last sampled prior to the occurrence.

In return for giving up total ordering and precise observation of the timing of 
events, we obtain the ability to employ push-based evaluation for event
occurrences, and the ability to non-deterministically merge event occurrences.
When events being input to a non-deterministic merge have simultaneous
occurrences, we arbitrarily select one to occur first. This does not violate any
guarantee about time values, since they will both have the same time value in
either case, and does not violate any guarantee about ordering, since no
guarantee of their order is given.

@@

In order to type signal functions, we must be able to describe their input and
output. In most signal function systems, a signal function takes exactly one
input signal and produces exactly one output signal. Multiple inputs or outputs
are handled by making the input or output a signal of tuples, and combinators
which combine or split the inputs or outputs of a signal assume this. Events are
represented at the type level as a particular type of signal, and at the value
level as an option: either an event occurrence or not.

@@

Our interface for evaluating signal functions takes the form of a monad
transformer~\cite{Jones1995}. This formulation permits us to provide specific
temporal actions to perform on a signal function in sequence, and in conjunction
with actions taken in Haskell's {\tt IO} monad, or another monad. Actions are
provided in this monad to push an event to the signal function's input, step
the signal function's time, or modify a signal on the signal functions input to
be observed at the next time step. Outputs are handled by a set of actuation
functions in the underlying monad, supplied by the user of the interface. This
set's type is parameterized over the signal function's output signal vector.

@@

TimeFlies includes basic signal functions and a library of signal function
combinators, grouped into lifting operations from pure functions to signal
functions, routing, reactivity, feedback, event processing, joining, and time
dependence.

These signal functions can be combined in serial (the output of one becomes
the input of the other) or in parallel (their input and output vectors are
combined). Signal functions are independently reactive or non-reactive, which
is the basis of their compositionality. Reactivity is introduced using the
{\tt switch} combinator, which takes an initial signal function, watches
one of its outputs for an event occurrence carrying a new signal function,
and replaces itself with this new signal function. Its type is:

\begin{code}
switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
          -> svIn :~> svOut
\end{code}

The full interface is described in the API documentation~\cite{TimefliesAPI}.

@@

{\tt SFEvalT} actions are constructed using combinators to push events,
update inputs, and step time, as well as actions lifted from the underlying
monad (used to obtain these inputs).

{\tt SFEvalT} actions are run using a supplied function which takes a signal
function's evaluation state and an {\tt SFEvalT} action and provides an action
in its underlying monad. This action includes the actions to actuate the outputs,
and returns the new evaluation state, along with the return value of the
{\tt SFEvalT} here.

Using this interface, the state can be stored in a mutable reference, or in a
closure. Handlers can be registered with the IO interface which push events, and
the actuation function can be constructed from the primitives given by the IO
library for actuation.

### 2012-11-30 Docs/IFL2012/discussion.tex
Performance results in discussion section - graphic not rendering properly

>>>>

We have described the implementation of a push-pull signal-function FRP system.
Push-based event evaluation should have advantages in event latency and overall
performance over pull-based event evaluation. We would also like to compare
the evaluation interface in the current widely-used system with the one proposed
here, again in order to verify that the theoretical benefits do appear in
practice.

One way to evaluate this is to write an application in two systems, and benchmark
these systems for comparison. In order to compare TimeFlies with Yampa, the most
recently released pull-based signal-function FRP system, we decided to implement
a suitable application in both.

\subsection{Evaluation Application}
\label{subsection:Evaluation_and_Comparisons-Evaluation_Application}

The application of choice is a learning switch controller for OpenFlow~\cite{OpenflowSpec}
switches. This application must handle input events triggered by switches
signalling that they have received a packet for which they have no handling
rules, and produce output events which install handling rules for dealing with
these packets.

The reactive algorithm is simple. A table from machine addresses to switch ports
is maintained. This table is updated in two ways. When a packet is sent to the
controller, meta-data for the packet indicates which port it arrived on. The
source address of the packet can then be mapped to that port in the table.
Once this occurs, rules are installed for each packet with a destination address
matching the source, along with reciprocal rules.

Each mapping in the table has a timestamp. After a certain amount of time, the
table entry is deleted. The table entry is also deleted if a packet is received
with the same source address and a different port. When an entry is deleted,
the switch is signaled to remove all corresponding rules.

In both Yampa and TimeFlies, this algorithm is implemented by using an accumulator,
which applies functions carried by event occurrences to a state value. It stores
the new state and outputs new event occurrences based on the output of the function.

The state stored is the table (implemented using the {\tt Data.Map} module in the
Haskell {\tt containers} library) and the output event occurrences carry messages
to be sent to the switch. The events are sourced internally to the learning
switch signal function (using the {\tt after} combinator in TimeFlies and the {\tt repeatedly}
function in Yampa) to delete expired rules, and externally, by transforming messages
from the switch to the controller into accumulator functions.

\subsection{Methodology and Results}
\label{subsection:Evaluation_and_Comparisons-Methodology_and_Results}

The controllers were benchmarked using the {\tt cbench} program~\cite{cbench}. The
amount to time to pause between time and signal updates (the inverse of the sampling rate for signals)
was passed as a parameter to both programs. The {\tt cbench} benchmark operates by sending a simulated packet in
event, waiting for a rule response, and then repeating. The output of the benchmark
is the number of times this occurs in a second. This is thus a very good test of
event latency in a reactive system.

Figure \ref{fig:timeflies-yampa-comparison} shows the performance comparison.
Toward the left of the graph is lower pause times (more frequent sampling). The
y-axis is a higher rate of responses (higher numbers are preferred).

\begin{figure}
\epsfig{file=graph.eps,angle=-90,width=\linewidth}
\hrule
\caption{Comparison of timeflies vs. Yampa implementing an OpenFlow learning switch controller.}
\label{fig:timeflies-yampa-comparison}
\end{figure}

There is some correlation between the sampling rate and the response rate
for TimeFlies, which is due to the continued wrapping of a replacement signal
function by {\tt switch} until the next time step (see Section~\ref{subsection:Implementation-Signal_Functions-Implementation_of_Signal_Function_Combinators}).
Nevertheless, TimeFlies outperforms Yampa in event response by several orders of magnitude,
even at a very high sampling rate. This is due to two factors. First, TimeFlies can react
to individual events without evaluating the whole signal function. Second, the
interface of time-flies permits us to connect its evaluation directly to GHC's event
manager, while the design of Yampa requires us to use threads and synchronized
communication\footnote{Yampa does include a step-wise evaluation function, but
it still couples event evaluation firmly to time steps, and is not well
documented.}.
